---
title: "Assignment 2: Risk Adjustment "
author: "Mathijs Lenderink"
date: '2022-11-06'
output: pdf_document
---

# Introduction
This analysis will show how much different variables influence the healthcare costs of an individual.
My findings in short:

* Increased age category has on average the result that you have more health costs.
* When you are male you will on average have more health costs.
* The different income sources have different sized effects on the health costs. 
* Whether or not you have limited coverage on your insurance has a large effect. When you have limited coverage on your insurance you have on average less health costs compared to someone who has full coverage.
* Whether or not you live in an unhealthy region also has a large effect on your health costs. When you live in an unhealthy region you have on average more health costs than someone who does not live in an unhealthy region.


# Assignment
Reading in libraries and data
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2); theme_set(theme_bw())
library(patchwork)
library(mlogit)
```

```{r}
data <- read.csv('data_assignment2.csv', sep = ',')
```

# Exploratory data analysis
```{r}
head(data)
```
Before I summarize the data I first set the categorical variables to categorical data type.
```{r}
categorical_cols <- c("Gender", "Age_category", "Insurer", "Income_source")
data[categorical_cols] <- lapply(data[categorical_cols], factor)
summary(data)
```
```{r}
dim(data)
```

The data summary shows that we have data of 996.308 people of which we know:

* ID: id of person
* Gender: gender of person (male/female)
* Age_category: in which age category the person falls, see below a summary of age categories and distribution of age.
* Order_age: the age_category ordered from low to high
* Insurer: Which insurer the person has
* Income_source: the source of income
* Limited_coverage: whether the person has limited coverage (yes/no)
* Unhealthy_region: whether the person lives in an unhealthy region (yes/no)
* Healthcare_cost: The healthcare cost 
* Population_density: how densely populated the area where the person lives is measured on a scale of 1 to 5

To check whether there are missing values:
```{r}
sapply(data, function(x) sum(is.na(x)))
```
There are no missing values.

For visualization and clarity purposes i set the age_category levels to increasing categories starting from category [0,5]
```{r}
age_levels <- c( "[0,5]", "(5,10]","(10,15]", "(15,20]", "(20,25]", "(25,30]", "(30,35]", "(35,40]", 
"(40,45]", "(45,50]",  "(50,55]", "(55,60]", "(60,65]", 
"(65,70]", "(70,75]", "(75,80]", "(80,85]", "(85,90]", "(90,95]", 
"(95,100]", "100+")
data$Age_category_leveled <- factor(data$Age_category, levels = age_levels)
```



# Data visualisation
## Basic graphs
```{r}
ggplot(data = data, aes( x = Age_category_leveled))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Distribution of Age")
```

```{r}
Gender_dist <- ggplot(data = data, aes( x = Gender))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Distribution of Gender")
```

```{r}
Income_dist <- ggplot(data = data, aes( x = Income_source))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Distribution of Income_source")
```



```{r}
Gender_dist + Income_dist
```
```{r}
Insurer_dist <- ggplot(data = data, aes( x = Insurer))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Distribution of people over insurers")
```

```{r}
mean_cost_insurer <- ggplot(data )+
  geom_bar(aes( x = Gender, y = Healthcare_cost ),stat = "summary", fun.y = "mean")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Distribution of average health cost epr isnurer")


```

```{r}
Insurer_dist + mean_cost_insurer
```
The above graphs show that insurer A,B,C,D all have the same average costs. However insurer E has a higher cost than the rest. This can be explained by the small market power (and number of insured) Insurer E has.

## Basic numbers

```{r}
data%>%
  group_by(Insurer)%>%
  summarise_at(vars(Limited_coverage), funs(mean(.)))
```
These results show that each insurer has around the same share of people who have limited coverage, all have around 7%.


```{r}
data%>%
  group_by(Insurer)%>%
  summarise_at(vars(Unhealthy_region), funs(mean(.)))
```
These results show that the number of people who live in an unhealthy region as a share per insurer is quite balanced. Each insurer has around 15% which live in an unhealthy region.


```{r}
data%>%
  group_by(Limited_coverage)%>%
  summarise_at(vars(Healthcare_cost), funs(mean(.)))
```
These results show that the mean healthcare costs of people with limited coverage are 0 and the mean healthcare costs of people without limited coverage is 8763.




## Exploratory Graphs
```{r}
ggplot(data = data, aes( x = Age_category_leveled, y = Healthcare_cost, color = Gender))+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Healthcare cost by age category and gender")
```
This boxplot shows that healthcare costs are increasing over age category and that overall the female healthcare costs are lower than the male healthcare costs. The below line graph shows the difference by gender.



The below graph shows that most of the people who have limited coverage are in the younger age groups. Up to the age group of 15 it can be expected that people do not have limited coverage as they probably fall under full coverage fo the government (such as the case in the Netherlands).
```{r}
ggplot(data = data, aes( x = Age_category_leveled, fill = as.factor(Limited_coverage)))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Distribution of Age by gender color by limited coverage")
```


The below graph shows that the distribution of people who are with a certain insurer does not change over age categories.
```{r}
ggplot(data = data, aes( x = Age_category_leveled, fill = as.factor(Insurer)))+
  geom_bar(position = "fill")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Percentage of insured per Insurer per Age category")
```

The Below graph shows that the percentage of people who liv ein an unhealthy region does not change over age categories.
```{r}
ggplot(data = data, aes( x = Age_category_leveled, fill = as.factor(Unhealthy_region)))+
  geom_bar(position = "fill")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Percentage of people who live in an unhealthy region per age category")
```





This graph shows that on average over all age groups the male healthcare costs are higher.
```{r}
ggplot(data = data, aes( x = Age_category_leveled, y = Healthcare_cost, colour = Gender))+
  stat_summary(aes(y = Healthcare_cost, group = Gender), fun.y = mean, geom = "line")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Healthcare cost by age category and gender")
```

This graph shows that people who live in an unhealthy region have on average per age category higher healthcare costs.
```{r}
ggplot(data = data, aes( x = Age_category_leveled, y = Healthcare_cost, colour = as.factor(Unhealthy_region)))+
  stat_summary(aes(y = Healthcare_cost, group = as.factor(Unhealthy_region)), fun.y = mean, geom = "line")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Healthcare cost by age category and Unhealthy region")
```

This graph shows that people with limited coverage have no healthcare costs.
```{r}
ggplot(data = data, aes( x = Age_category_leveled, y = Healthcare_cost, colour = as.factor(Limited_coverage)))+
  stat_summary(aes(y = Healthcare_cost, group = as.factor(Limited_coverage)), fun.y = mean, geom = "line")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Healthcare cost by age category and limited coverage")
```




```{r}
ggplot(data = data, aes( x = Age_category_leveled, fill = Gender))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Distribution of Age by gender")
```
In the graph above we can see that the share of males in the youngest categories is very large, also in the oldest categories this difference can be observed.

# Estimating model based on Age and Gender

```{r}
model1 <- lm(Healthcare_cost ~ Age_category_leveled + Gender, data = data)
summary(model1)

```
The simple model with age and gender above shows that, just like in the graphs, the costly individuals are older individuals and males in general. Males have on average 1780 more healthcost. The older the individual the more healthcosts you will on average have, this can be observed from the increasing coefficient of the age_categories. The older the category the higher the coefficient estimate of the age_category, meaning that on average an individual will have higher health costs when they fall in a higher age category.
```{r}
age_coefficients <- model1$coefficients[2:21]
age_index <- seq(20)
data_age_coeff <- data.frame(age_index, age_coefficients)
ggplot(data_age_coeff, aes(x = age_index, y = age_coefficients))+
  geom_line()+
  ggtitle("Estimated coefficients per age category")
```

## SImple model with split gender per age category

Below I create a model where each age category is also split by gender to see whether the gender costs coefficietn change over age categories. (Here I set the intercept to 0 to make the coefficients positive)

```{r}
model_Split_gender <- lm(Healthcare_cost ~ (Age_category_leveled:Gender) -1, data = data)
summary(model_Split_gender)

```

The coefficients show that there is a difference between genders and that there is an increasing trend over age categories, meaning that just as in the graph above the health costs are increasing the older someone is. To make visible the difference between gender per age category I again plotted the coefficients below, but then split per gender. The graph shows the same trend as the earlier graph on average health costs shows above (see the graph with title: Healthcare cost by age category and gender).

```{r}
age_coefficients_female <- model_Split_gender$coefficients[2:21]
age_coefficients_male <- model_Split_gender$coefficients[23:42]
age_index <- seq(20)
data_age_coeff_gender <- data.frame(age_index, age_coefficients_female, age_coefficients_male)
data_age_coeff_gender_long <- pivot_longer(data_age_coeff_gender, cols = c(age_coefficients_female, age_coefficients_male), names_to = "Source", values_to = "Value")

ggplot(data_age_coeff_gender_long, aes(x = age_index, y = Value, color = Source))+
  geom_line()+
  ggtitle("Estimated coefficients per age category per gender")
```


## A second model based on the ordered age:
Order_age is a variable ranging from 1 to 24 depending on the age category, the higher the number the higher the age category. a one increase in the Order_age means one higher age category. For most of the data (except above age 100) this means that a person is in an age class of 5 years higher. See below for a table of Order_age values per Age category.
```{r}
age_table <- data %>%
  group_by(Order_age)%>%
  distinct(Age_category)
colnames(age_table) <- c("Age_category", "age_index")
age_table$age_index = age_table$age_index - 1
age_table <- age_table[0:21,]
age_table
```


```{r}
model2 <- lm(Healthcare_cost ~ Order_age + Gender, data = data)
summary(model2)
```
The outcome of this regression shows that on average when you increase Order_age by 1, so fall in an age category higher, you will have 1032.6 increased health costs.
The outcome from this regression also shows that you will on average have 926.8 increased health costs when you are male instead of female.


To conclude from both regressions the groups that are likely to be profitable and the groups that are likely to be loss-making:

* Profitable:
  * Females on average have lower healthcare costs.
  * The older people are the more the healthcare costs. To get insight in how this changes over time I have plotted the coefficients in the graph above. There is a large jump between age category [25-30] with a value 2737 and age category [30-35] with a value 6141. See the graph for the exact change in coefficient. But it depends on the premium of the individuals at what age the individuals become loss-making on average.

* Loss-making:
  * Males on average have higher healthcare costs.
  * Older individuals have increased healthcare costs (see the coefficients per age category graph above). Older people are more likely to be loss making.


# Model Extension

I will now extend the model using other available data and analyze whether this extra data increases the accuracy of the model.
```{r}
summary(model1)
```

```{r}
model3 <- lm(Healthcare_cost ~ Age_category_leveled + Gender + Income_source + Limited_coverage + Unhealthy_region + Population_density, data = data)
summary(model3)
```
Population density will be left out of the model as the estimated effect size is very small and the coefficient is not significant. The other variables I will leave in the model, in the appendix summaries can be found on the models with and without the other variables. From those I conclude that model fit (R-squared) and statistical significance are optimal when I leave in all the variables except population density. 
The other added variables do seem to have a significant effect on the health costs based on the estimated coefficients and the statistical significance of these coefficients. 
Another important measure to check whether the model has become more accurate with the added variables is the value of the R-squared of the model.
Compared to the model without the added variables we see an increase in the R-squared.
R-squared value of the basic age and gender model: 0.6076
R-squared value of the model with added variables: 0.6852


When we remove the population density variable from the model we are left with the following model:
```{r}
model4 <- lm(Healthcare_cost ~ Age_category_leveled + Gender + Income_source + Limited_coverage + Unhealthy_region , data = data)
summary(model4)
```
Compared to the previous model the R-squared has not changed, which also indicates that the population density did not add accuracy to the model.


## Model analysis
From the estimated coefficients of the final model we can observe the following:

* Increased age category has on average the result that you have more health costs.
* When you are male you will on average have 1785 more health costs.
* The different income sources have different sized effects on the health costs. It can be expected that someone who has unemployment benefits have on average a higher health cost than someone who works and it can be expected that students (often young and healthy) will have lower health costs than people who live on a pension. This reasoning can be seen in the estimated coefficients:
  * Income_sourcePension:                1445.114 
  * Income_sourceStudent:                1353.337 
  * Income_sourceUnemployment Benefits:  1430.392 
  * Income_sourceWorking:                1365.002 
* Whether or not you have limited coverage on your insurance has a large effect. When you have limited coverage on your insurance you have on average 3909.3 less health costs compared to someone who has full coverage.
* Whether or not you live in an unhealthy region also has a large effect on your health costs. When you live in an unhealthy region you have on average 3862.0 more health costs than someone who does not live in an unhealthy region.



# Further analysis using graphs
As we concluded above the individuals who are in the category [30-35] and above have a much higher health cost than those younger than them. So for analysis I split this group in two where one group is everyone under 30 and one group is everyone above 30. In the grpah below this shows that I will include everyone up to the green line.

```{r}
data$IsYoung <- ifelse(data$Order_age <= 6, 1, 0)
```


```{r}
ggplot(data_age_coeff, aes(x = age_index, y = age_coefficients))+
  geom_line()+
  ggtitle("Estimated coefficients per age category")+
  geom_vline(xintercept = 5.5, colour = 'green')

ggplot(data = data, aes( x = Age_category_leveled, fill = as.factor(IsYoung)))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Distribution of Age")



```

Using this split I can now visualize which insurer has the most profitable individuals.

```{r}
ggplot(data = data, aes( x = Insurer, fill = as.factor(IsYoung)))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Distribution of people over the insurers")
```


```{r}
data%>%
  group_by(Insurer)%>%
  summarise_at(vars(IsYoung), funs(mean(.)))
```

From this analysis we can observe that Insurer E has the highest share of people under 30, namely 23.9%. However the percentages between insurers do not differ much. The lowest percentage is 23.67%, while the highest (from insurer E) is 23.91%


# Appendix
```{r}
model_A1 <- lm(Healthcare_cost ~ Age_category_leveled + Gender  + Limited_coverage + Unhealthy_region + Population_density, data = data)
summary(model_A1)
```
```{r}
model_A2 <- lm(Healthcare_cost ~ Age_category_leveled + Gender + Income_source  + Unhealthy_region + Population_density, data = data)
summary(model_A2)
```
```{r}
model_A3 <- lm(Healthcare_cost ~ Age_category_leveled + Gender + Income_source + Limited_coverage  + Population_density, data = data)
summary(model_A3)
```
```{r}
model_A4 <- lm(Healthcare_cost ~ Age_category_leveled + Gender + Income_source + Limited_coverage + Unhealthy_region , data = data)
summary(model_A4)
```